---
title: "TabulaMurisSenisData"
author: "Charlotte Soneson, Dania Machlab, Federico Marini"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('TabulaMurisSenisData')`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{TabulaMurisSenisData}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{TabulaMurisSenisData}
  %\VignetteKeywords{GeneExpression, RNASeq, SingleCell}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: TabulaMurisSenisData.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(SingleCellExperiment)
library(TabulaMurisSenisData)
```

# Introduction

This package provides access to the processed bulk and single-cell RNA-seq data 
from the _Tabula Muris Senis_ data set 
[@Schaum2019-nf;@Tabula_Muris_Consortium2020-um]. The processed bulk RNA-seq 
data was downloaded from GEO (accession number 
[GSE132040](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE132040)). The 
single-cell data (droplet and FACS) was obtained from FigShare (see below for links).
All data sets are provided as `SingleCellExperiment` objects for easy access 
and use within the Bioconductor ecosystem. 

# Bulk RNA-seq data

```{r}
tms_bulk <- TabulaMurisSenisBulk()
tms_bulk
```

# Single-cell RNA-seq data

## Droplet data

The data files for the droplet single-cell RNA-seq data were downloaded from FigShare: 

* [The h5ad file with the raw counts](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23938934)
* [The h5ad file with the processed counts, and additional reduced dimensionalities](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23936684)


```{r}
tms_droplet <- TabulaMurisSenisDroplet()
tms_droplet
```

## FACS data

The data files for the FACS single-cell RNA-seq data were downloaded from FigShare: 

* [The h5ad file with the raw counts](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23939711)
* [The h5ad file with the processed counts, and additional reduced dimensionalities](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23937842)

```{r}
tms_facs <- TabulaMurisSenisFACS()
tms_facs
```

# Session info

```{r}
sessionInfo()
```

# References
