---
title: "Download and preprocess the Tabula Muris Senis facs data"
author: "Dania Machlab"
date: "2020-12-03"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 100)
```

# Loading required packages


```{r libraries}
suppressPackageStartupMessages({
    library(utils)
    library(HDF5Array)
    library(SingleCellExperiment)
})
```


# Downloading the `h5ad` files from figshare

The following files were downloaded on Dec 3, 2020:

[The h5ad file with the raw counts](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23939711)

[The h5ad file with the processed counts, and additional reduced dimensionalities](https://figshare.com/articles/dataset/Processed_files_to_use_with_scanpy_/8273102?file=23937842)

```{r download, eval=FALSE}
## Create data dirs
dir.create("facs-raw-data")
dir.create("facs-raw-data/raw", showWarnings = FALSE, recursive = TRUE)
dir.create("facs-raw-data/processed", showWarnings = FALSE, recursive = TRUE)
dir.create("TabulaMurisSenisData/tabula-muris-senis-facs", showWarnings = FALSE, recursive = TRUE)

## Download data (set timeout time to 360 sec, default is 60 sec)
url_raw="https://ndownloader.figshare.com/files/23939711"
url_processed="https://ndownloader.figshare.com/files/23937842"

options(timeout=360)
download.file(url = url_raw, destfile = file.path("facs-raw-data", "raw", "raw.h5ad"))
download.file(url = url_processed, destfile = file.path("facs-raw-data", "processed", "processed.h5ad"))
```

# Saving as `loom` and `csv` files in `python`

We save the count matrices as loom files, and the metadata as csv files in `python`. The reason for this is that due to version incompatibilities, we cannot directly load the `h5ad` files into `R`.

```{python eval=FALSE}
import scanpy as sc

## Read h5ad files
ad_raw=sc.read_h5ad("facs-raw-data/raw/raw.h5ad")
ad_processed=sc.read_h5ad("facs-raw-data/processed/processed.h5ad")

## Save rowData, colData, and metadata as csv files
# ... obs.csv is our colData
# ... var.csv is our rowData
# ... obsm.csv is : 1) the PCA (50 PCs) on the cells from col 1:50
# ... ...           2) the UMAP on the cells in cols 51 and 52
ad_raw.write_csvs("facs-raw-data/raw")
ad_processed.write_csvs("facs-raw-data/processed")

## Save count matrices as loom files
ad_raw.write_loom("facs-raw-data/raw/raw.loom", write_obsm_varm=False)
ad_processed.write_loom("facs-raw-data/processed/processed.loom", write_obsm_varm=False)
```

# Processing and saving in `R`

We read in the count matrices as delayed arrays, and save as HDF5 files. We read in the metadata files, organize them, and save them as `csv` files.

```{r counts, eval=FALSE}
## Read in as DelayedArray
counts_raw <- HDF5Array::HDF5Array(filepath = "facs-raw-data/raw/raw.loom", name = "matrix")
counts_processed <- HDF5Array::HDF5Array(filepath = "facs-raw-data/processed/processed.loom", name = "matrix")

## transpose to have genes x cells
counts_raw <- t(counts_raw)
counts_processed <- t(counts_processed)

## Save (takes ~ 20-30 min per matrix)
options(DelayedArray.block.size=1e9) # 1GB block size.
mat_raw <- writeHDF5Array(x = counts_raw, 
                          filepath = file.path("TabulaMurisSenisData", "tabula-muris-senis-facs", "counts.h5"), 
                          name = "counts", 
                          chunkdim = HDF5Array::getHDF5DumpChunkDim(dim(counts_raw)))

mat_processed <- writeHDF5Array(x = counts_processed, 
                                filepath = file.path("TabulaMurisSenisData", "tabula-muris-senis-facs", "processed_counts.h5"), 
                                name = "processed_counts", 
                                chunkdim = HDF5Array::getHDF5DumpChunkDim(dim(counts_processed)))
```

The raw and processed h5ad files from figshare have the cells and genes in the same order. The rawData and colData matrices are identical, except the processed one has some additional columns, which is why we include them.

```{r rowData}
## Read in csv
rowData <- utils::read.table("facs-raw-data/processed/var.csv", sep = ",", header = TRUE)

## Save
saveRDS(rowData, "TabulaMurisSenisData/tabula-muris-senis-facs/rowData.rds")
```

```{r colData}
## Read in csv
colData <- utils::read.table("facs-raw-data/processed/obs.csv", sep = ",", header = TRUE)

## add tissue cols as in (https://tabula-muris-senis.ds.czbiohub.org)
tissue_cols <- read.csv("facs-raw-data/processed/uns/tissue_colors.csv", header = FALSE)[,1]
names(tissue_cols) <- unique(colData$tissue)
tissue_cols
m <- match(colData$tissue, names(tissue_cols))
colData$tissue_col <- tissue_cols[as.factor(colData$tissue)]

## Save
saveRDS(colData, "TabulaMurisSenisData/tabula-muris-senis-facs/colData.rds")
```

We read in the PCA and UMAP from the processed data and save them as `rds` files.

```{r reduced_dims}
## Read csv
pca_and_umap <- read.table("facs-raw-data/processed/obsm.csv", sep=",", header = TRUE)

## Extract matrices
umap <- as.matrix(pca_and_umap[, grep("umap", colnames(pca_and_umap), ignore.case = TRUE)])
pca <- as.matrix(pca_and_umap[, grep("pc", colnames(pca_and_umap), ignore.case = TRUE)])

## Save
saveRDS(umap, "TabulaMurisSenisData/tabula-muris-senis-facs/umap.rds")
saveRDS(pca, "TabulaMurisSenisData/tabula-muris-senis-facs/pca.rds")
```

# Session

```{r session}
date()
sessionInfo()
```

